---
title: "Creating biodiversity projections"
author: "Martin Jung"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating biodiversity projections}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Define variables for vignette figures and code execution
h <- 5.5
w <- 5.5
is_check <- ("CheckExEnv" %in% search()) || any(c("_R_CHECK_TIMINGS_",
             "_R_CHECK_LICENSE_") %in% names(Sys.getenv()))
knitr::opts_chunk$set(fig.align = "center", eval = !is_check)
```

Besides inferring environmental responses and mapping the contemporary distribution of any species (see [Train simple Model](01_train_simple_model.html)), species distribution models (SDMs) are commonly used to make predictions on how biodiversity is expected to change under future conditions. Usually this is being done by first training a model on present observations and then projecting the resulting $\beta$ coefficients to another set of environmental predictors from another time period or spatial region.  

The `ibis.iSDM` R-package provides direct support for creating such projections. This requires a previously fitted [`DistributionModel`] and a new set of covariates which match in names the covariates on which the fitted model was trained. The key functions here are `scenario()` and `project()` which are specific to such projections and result in the creation of a [`BiodiversityScenario`] object. For other functions the same syntax as for previously trained models applies, e.g. new predictor can be added via `add_predictors()` or thresholds applied via `threshold()`. In addition the `ibis.iSDM` R-package allows the specification of a number of constrains on the projections, such as for instance dispersal constrains based on the expected or simulated chance of individuals dispersing to neighbouring grid cells. Such constrains can be added via `add_constrain()` (and other constrain functionalities in `add_constrain_*()`).

**Note**:
Projecting relative suitability estimates into conditions other than the observational records for
which the SDM was trained comes with a number of assumptions, most importantly that relationships between occurrences and environmental conditions are in equilibrium and similar biases and conditions apply in the future. See [Elith et al. (2010)](https://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2010.00036.x) and discussion in [Zurell et al. (2016)](http://doi.wiley.com/10.1111/gcb.13251) for an introduction and comparative overview.

```{r eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
# Install for this vignette if not present
# Function to check for installed packages and install them if they are not installed
install <- function(packages){
  new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
  if (length(new.packages)) 
    install.packages(new.packages, dependencies = TRUE)
  sapply(packages, require, character.only = TRUE)
}
invisible( install(c("xgboost", "igraph", "rgeos", "kissmig", "stars", "ncdf4","geosphere", "ggplot2","cubelyr")) )
```

## Load relevant packages and testing data
```{r Load the package, message=FALSE, eval=TRUE}
# Load the packages
library(ibis.iSDM)
library(stars)
library(xgboost)
library(raster)
library(igraph)
library(ggplot2)
library(ncdf4)
library(assertthat)
```

For the purpose of this example we are loading some testing data on species distributions as well as contemporary and future predictors. Note the names of predictors used for building a distribution model have to be consistent with those for creating projections!

```{r Load data, warning=FALSE, eval = TRUE, fig.width = w, fig.height = h}
# Background and biodiversity data
background <- raster::raster(system.file('extdata/europegrid_50km.tif', package='ibis.iSDM'))
virtual_points <- sf::st_read(system.file('extdata/input_data.gpkg', package='ibis.iSDM'), 'points', quiet = TRUE)

# Note we are loading different predictors than in previous examples
# These are in netcdf4 format, a format specific for storing spatial-temporal data including metadata.
ll <- list.files(system.file("extdata/predictors_presfuture/", package = "ibis.iSDM", mustWork = TRUE), "*.nc",full.names = TRUE)

# From those list of predictors are first loading the current ones as raster data
# We are loading only data from the very first, contemporary time step for model fitting
pred_current <- raster::stack()
for(i in ll) pred_current <- addLayer(pred_current, raster::raster(i, layer = 1) )
projection(pred_current) <- projection(background) # Same projection as background

# Get future predictors
# These we will load in using the stars package and also ignoring the first time step
pred_future <- stars::read_stars(ll) |> stars:::slice.stars('Time', 2:86)
st_crs(pred_future) <- st_crs(4326) # Set projection
# Rename future predictors to those of current
names(pred_future) <- names(pred_current)

# Plot the test data
plot(pred_current$Secondary.forest.vegetation,
     col = colorRampPalette(c("grey20", "orange", "lightgreen", "green"))(10),
     main = "Share of secondary vegetation")
```
## Train model and create a future projection

We will make use of the data loaded above to (a) first create a species distribution model for contemporary conditions and (b) project the obtained coefficients into the future using future predictors. 
For guidance on how distribution models are trained, see other vignettes ([1](01_train_simple_model.html)).

```{r Train a basic model, eval = TRUE}
# Train model adding the data loaded above
x <- distribution(background) |> 
  add_biodiversity_poipo(virtual_points, field_occurrence = 'Observed', name = 'Virtual points') |> 
  # Note that we scale the predictors here
  add_predictors(pred_current, transform = 'scale',derivates = 'none') |> 
  engine_glmnet(alpha = 0) 

# Train the model
modf <- train(x, runname = 'Simple PPM', verbose = FALSE)

# Add a threshold to this model by getting 05 percentile of values
modf <- threshold(modf, method = 'percentile', value = 0.05)

# -- #
# Now lets create a scenarios object via scenarios
sc <- scenario(modf) |> 
  # Apply the same variable transformations as above. 
  add_predictors(pred_future, transform = 'scale') |> 
  # Calculate thresholds at each time step. The threshold estimate is taken from the model object.
  threshold()

# This creates a scenario object
sc
# The object contains its own functions. See the scenarios help file for more information on 
# what is possible with them
names(sc)

```
The scenario object can finally be trained via `project()`.

```{r Create the scenario}
sc.fit1 <- sc |> project()
# Note that an indication of fitted scenarios has been added to the object
sc.fit1
```

## Summarizing and plotting the fitted projections

As with distribution models there are a number of ways how the scenarios can be visualized 
and interacted with:

* `plot()` makes a visualization of the projections over all time steps (!)
* `plot_relative_change()` calculates the change in suitability area between the first and the last timestep and categorizes the result accordingly. Note that SDMs as such cannot directly infer colonization or extinction, but only
gains or losses of suitable habitat!
* `calc_scenarios_slope()` calculates the slope (rate of change) across timesteps. Useful for summarizing results
* `summary()` creates a summary output of the contained scenarios. If a `threshold()` is specified, this function
will summarize the amount of area at each timestep.
* `get_data()` gets the created scenarios a `stars` object (plus *thresholds* if specified).


```{r Plotting and summarizing the created projections, eval = TRUE, fig.width = w, fig.height = h}
# Plot all scenarios. With a large number of predictors this figure will be messy...
plot(sc.fit1) # or sc.fit1$plot()

# As an alternative, visualize the linear slope per grid cell and across all time steps
o <- sc.fit1$calc_scenarios_slope()
plot(o)

# Another option is to calculate the relative change between start and finish
o <- sc.fit1$plot_relative_change(plot = TRUE)
print(o)

# We can also summarize the thresholded data
o <- sc.fit1$summary()
plot(area_km2~band, data = o, type = 'b',
     main = "Suitable habitat across Time",
     ylab = "Amount of area (km2)", xlab = "Time")

# How does habitat gain and loss change over time?
plot(totchange_gain_km2~band, data = o, type = 'n',
     main = "Habitat gain and loss", ylim = c(-2e4, 2e4),
     ylab = "Amount of area (km2)", xlab = "Time")
lines(o$totchange_gain_km2~o$band, col = "blue")
lines((o$totchange_loss_km2)~o$band, col = "red")

```

Finally, scenarios projections can also be saved as specific outputs. As before, this is enabled via
`write_output()` and works just the same for [`BiodiversityScenario`] objects, with the only difference being that the output can be specified as netCDF-4 file.

## Adding constraints to projections

In the simple scenario above we use the naive assumption that any, depending on the response functions of the fitted distribution model, any suitable habitat within the background modelling region is potentially reachable by the species. In reality there might however be geographic (e.g. islands), environmental and biotic constraints on how far a species can disperse. These can be specified through the constrain function [`add_constraint()`] and a variety of constraints is currently available, some of which depend on other packages.

* `add_constraint()` Generic wrapper to which a specific 'method' can be supplied. See documentation for more information on available options and parameters.
* `add_constraint_dispersal()` To add a dispersal constraint to the projections which is applied after each time step. Supports various options with `'sdd_fixed'` for fixed dispersal kernels, `'sdd_nexpkernel'` for a negative exponential kernel or `'sdd_kissmig'` for applying the kissmig framework.
* `add_constraint_MigClim()` Use the MigClim R-package to simulate dispersal events between time steps. A number of parameters are required here and adding this constrain will also overwrite some default plotting capacities (For example via `sc$plot_migclim()`). See also the help file and [Engler et al. (2012)](https://onlinelibrary.wiley.com/doi/10.1111/j.1600-0587.2012.07608.x) for more information.
* `add_constraint_connectivity()` Add a connectivity constrain to the projection. Currently only hard barriers are implemented, but in future additional sub-modules are planned to enable more options here.
* `add_constraint_adaptability()` Simple constraints on the adaptability of species to novel climatic conditions. Currently only simple nichelimits are implemented, which 'cap' projections in novel environments to the observed ranges of contemporary predictors.
* `add_constraint_boundary()` Specifying a hard boundary constraint on all projections, for example by limiting (future) projections to only a certain area such as a biome or a contemporary range.

Lastly there are also options to *stabilize* suitability projections via the `project()` function. Specifying a stabilization here results in the projections being smoothed and informed of incremental time steps. This can particularly help for projections that use variables known to make sudden, abrupt jumps between time steps (e.g. precipitation anomalies).  

```{r Add constraints and reproject, eval = TRUE, fig.width = w, fig.height = h}
# Adding a simple negative exponential kernel to constrain the predictions
sc.fit2 <- sc |>   
   add_constraint(method = "sdd_nex", value = 1e5) |> 
   # Directly fit the object
   project(stabilize = F)

# Also fit one projection a nichelimit has been added
sc.fit3 <- sc |>   
   add_constraint(method = "sdd_nex", value = 1e5) |> 
   add_constraint_adaptability(method = "nichelimit") |> 
   # Directly fit the object
   project(stabilize = F)

# Note how constrains are indicated in the scenario object.
sc.fit3

# The naive assumption is that there is unlimited dispersal across the whole background
# Note how the projection with dispersal constrain results in a considerable smaller amount of suitable habitat.
sc.fit1$plot(which = 40) # Baseline
sc.fit2$plot(which = 40) # With dispersal constrain
sc.fit3$plot(which = 40) # With dispersal limit and nichelimitation (within a standard deviation)

```

```{r Summarize area of projections, eval = TRUE, fig.width = w, fig.height = h}
# Lets compare the difference in projections compared to the naive one defined earlier. 
o1 <- sc.fit1$summary()
o2 <- sc.fit2$summary()
o3 <- sc.fit3$summary()
arlim <- c(min(o1$area_km2, o2$area_km2, o3$area_km2)-5000,
           max(o1$area_km2, o2$area_km2, o3$area_km2))

plot(area_km2~band, data = o1, type = 'n',
     ylim = arlim,
     main = "Suitable habitat projection",
     ylab = "Amount of area (km2)", xlab = "Time")
lines(o1$area_km2~o1$band, col = "black", lty = 1)
lines(o2$area_km2~o2$band, col = "black", lty = 2)
lines(o3$area_km2~o3$band, col = "black", lty = 3)
legend("bottomleft", 
  legend = c("Unlimited dispersal", "Constrained dispersal",
             "Constrained dispersal and niche limit"), 
  lty = c(1, 2, 3),
  cex = 1.2,
  bty = "n")

# Lastly it is also possible to directly summarize the state 
# before (usually first year) and end (last year).
sc.fit2$summary_beforeafter()

```

Another option for constraining prediction is also by imposing a zonal limit (for instance climatically defined) on the projections (see alternatively `add_constraint_boundary()` above). This has to be done while fitting the SDM for the reference conditions (see the example with limits ([1](01_train_simple_model.html)) ) and is considered when doing (future) projections.

## Specific parsers for GLOBIOM related scenarios

IIASA's Global Biosphere Management Model ([GLOBIOM](https://iiasa.github.io/GLOBIOM/)) is a partial equilibrium model and used to analyze the competition for land use between agriculture, forestry, and bioenergy, which are the main land-based production sectors. It builds on.

With the ibis.iSDM being part of IIASA's suite of integrated models, there is a direct link available to make use of downscaled GLOBIOM outputs. Implemented are functions to either directly format the data via [`formatGLOBIOM()`] or add them to any `DistributionModel-class` or `BiodiversityScenario-class` object directly via `add_predictors_globiom()` . 
